---
title: "Analysis of acute theatre usage"
author: "Dr Greig Russell"
date: "7 February 2018"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

#Acute theatre usage
The aim of this analysis was to consider usage of acute theatres. The original intention was for the analysis to focus on only Nov.2017, but the data was not initially available. The first phase of the analysis therefore focused on June.2017. Subsequently the Nov.2017 data became available, allowing the analysis to be expanding. Comparison of utilization for acute cases between the two months reveals some unexpected differences. 

##Executive summary
The results from June describe how most acute cases are done in the acute theatre. The short duration cases are done at night, whilst the longer duration acute cases are done during working hours. The ratio of acute cases undertaken during the day compared to after hours is 2:1. By specialty General Surgery is the most prevalent followed by orthopedic surgery and the O&G. Cases build by over the week, making Friday the most common day for acute cases to occur, with this build up being most marked in General Surgery. 

In contrast November 2017 reveals a number of changes. The number of acute cases has reduced in general, but especially in the after hours. Notably, but not significantly, O&G has reduced its utilization of acute theatre in Nov.2017. Associated with this reduction in utilization was a statistically significant growth in the mean duration of cases. There is a slight shift back in the busiest day of acutes from Friday in June to Thursday in November. The shift is slight, not significant and is accompanied by more orthopaedic acutes in November as compared to General surgery. The number of cases performed in each theatre has remained similar.

Given the later I performed a separate analysis of "acute arranged"" to test if the reduction was a classification artificial, but this analysis did not reveal significant change between June and November. If anything it mirrored the changes seen in the Acute theatre utilization.

The most likely cause for the changes seen in this analysis is a cultural shift so that acute surgery, particular after hours, to only those cases where life or limb are under threat. 

```{r load_the_libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(knitr)
library(rmarkdown)
library(gridExtra)
```

```{r load_the_data, echo=FALSE, message=FALSE, warning=FALSE}
Column_names <- c(
  "Counter",
  "NHI",
  "Theatre",
  "Type",
  "Original_Ward",
  "Surgeon",
  "Surgeon_Teach",
  "Anaesthetist",
  "Anaesthetist_Teach",
  "Anaesthetic",
  "Anaesthetist_Risk",
  "Operation_Code",
  "Operation_Code_Extention",
  "PurchaserNumber",
  "Patient_Side",
  "Specialty_Over_ride",
  "Booking_In_Date",
  "Booking_In_Time",
  "Event_Date",
  "Arrival_In_Anaesthetic_Room",
  "Anaesthesia_Start",
  "Patient_Ready_for_Operation",
  "Operation_Start",
  "Operation_End",
  "Recovery_Takeover",
  "Patient_Opens_Eyes_On_Command",
  "Recovery_Complete",
  "Left_Recovery",
  "Anaesthetic_Time",
  "Operation_Time",
  "Scrub_Nurse",
  "Charge_Nurse",
  "Anaesthetic_Assist",
  "AnaestheticAssistTeach",
  "Problem_Code",
  "DirectMaterialCost",
  "Planned_Return",
  "SessionCode",
  "Verified",
  "Episode",
  "AdmissionType",
  "Modified",
  "ModifiedBy",
  "Processed",
  "Created",
  "EventComments",
  "TheatreOther",
  "PlannedOperation",
  "GRN",
  "Admission_Date",
  "AdmissionWard",
  "EpisodeWhole",
  "TimeOut",
  "UnplannedReturn",
  "Problem_Description",
  "Group_Description",
  "Specialty_Description",
  "SpecialtyCode",
  "Operation_Description"
)

OTIS <- read_csv("/home/greig/R-projects/Clinical_audits/Theatre/OTIS_rawData.csv", col_names = Column_names, skip = 1)
```

```{r tidy_the_data, echo=FALSE, message=FALSE, warning=FALSE}
OTIS$Operation_Start <- as.character(OTIS$Operation_Start)
OTIS$Operation_End <- as.character(OTIS$Operation_End)

OTIS$Operation_Start <- paste(OTIS$Event_Date, OTIS$Operation_Start, sep = " ")
OTIS$Operation_End <- paste(OTIS$Event_Date, OTIS$Operation_End, sep = " ")
OTIS$Anaesthesia_Start <- paste(OTIS$Event_Date, OTIS$Anaesthesia_Start, sep = " ")
OTIS$Recovery_Takeover <- paste(OTIS$Event_Date, OTIS$Recovery_Takeover, sep = " ")

OTIS$Operation_Start <-  dmy_hms(OTIS$Operation_Start)
OTIS$Operation_End <- dmy_hms(OTIS$Operation_End)
OTIS$Anaesthesia_Start <- dmy_hms(OTIS$Anaesthesia_Start)
OTIS$Recovery_Takeover <- dmy_hms(OTIS$Recovery_Takeover)

OTIS$Surg_Span <- interval(OTIS$Operation_Start, OTIS$Operation_End)
OTIS$Surg_Span <- as.duration(OTIS$Surg_Span)
OTIS$Surg_Span <- as.integer(OTIS$Surg_Span)  / 60

OTIS$Anaes_Span <- interval(OTIS$Anaesthesia_Start, OTIS$Recovery_Takeover)
OTIS$Anaes_Span <- as.duration(OTIS$Anaes_Span)
OTIS$Anaes_Span <- as.integer(OTIS$Anaes_Span)  / 60

OTIS$Event_Date <- dmy(OTIS$Event_Date)

OTIS <- OTIS %>% 
  filter(0 < Surg_Span) %>% 
  filter(0 < Anaes_Span)

OTIS$AdmissionType <- as.factor(OTIS$AdmissionType)

OTIS$DOW <- wday(OTIS$Event_Date)
OTIS$Hour <- hour(OTIS$Anaesthesia_Start)
OTIS$Theatre <- as.factor(OTIS$Theatre)

OTIS$TOD <-  ifelse(8 <= OTIS$Hour & OTIS$Hour <= 17, "Daytime", "Afterhours" )
OTIS$TOD <- as.factor(OTIS$TOD)

OTIS$Operation_Description <- as.factor(OTIS$Operation_Description)

AcuteOT <- OTIS %>% 
  filter("2017-06-01" <= Event_Date & Event_Date <= "2017-06-30" & AdmissionType == "AC") %>% 
  filter(Theatre != "M" & Theatre != "O") %>% 
  filter(Anaes_Span < 400)
```

##June.2017 data

###Theatre location
As expected the predominance of acute cases in June.2017 occurred in Theatre Four, the designated acute theatre.

```{r june_acute_location, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = AcuteOT) +
  geom_bar(aes(x = Theatre, fill = Theatre)) +
  labs(title = "Number of cases per theatre June.2017", x = "Theatre", y = "Count of Acute cases") +
  theme_minimal()
```

###Hour of the day
Most commonly acute cases occurred during the day (141 cases) s compared to after hours (62 cases). 

```{r june_tod, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = AcuteOT, aes(x = Hour)) +
  geom_bar(aes(fill = TOD)) +
  labs(title = "Number of cases by hour of the day June.2017") +
  guides(fill = guide_legend(title = "Time of Day")) +
  theme_minimal()
```

Which is more easy to visualize using a probability density plot

```{r june_density_tod, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = AcuteOT, aes(x = Hour)) +
  geom_density(aes(fill = TOD)) +
  labs(title = "Density plot by hour of the day June.2017") +
  guides(fill = guide_legend(title = "Time of Day")) +
  theme_minimal()
```

###Day of week
Unexpectedly acute cases became more prevalent as the week proceeded, with the lowest volume of cases being in the weekend.

```{r june_dow, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = AcuteOT) +
  geom_bar(aes(x = DOW, fill = factor(DOW, labels = c("Sun", "Mon", "Tues", "Wed", "Thur", "Fri", "Sat")))) +
  labs(title = "Count of cases by Day of Week, June.2017", x = "Day of week") +
  guides(fill = guide_legend(title = "Day of week")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks = element_blank())
```

###Case duration
The cases performed after hours were of shorter duration, than those completed during working hours.

```{r june_duration, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(AcuteOT) +
  geom_density(aes(Anaes_Span, fill = TOD)) +
  labs(title = "Density plot by duration & time of the day", x = "Anaesthetic duration (minutes)") +
  guides(fill = guide_legend(title = "Time of Day")) +
  theme_minimal()
```

###Surgical sub-speciality
Possibly anticipated but General Surgery performed the most acute cases, followed by Orthopaedic surgery and then O&G.

```{r june_subspec, echo=FALSE, warning=FALSE, message=FALSE}
SubSpec <- AcuteOT %>% 
  group_by(Specialty_Description) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

kable(SubSpec)
```

When considering the distribution of cases across the week by surgical subspeciality (and where 1 = Sunday), the results were;

```{r june_subspec_day, echo=FALSE, warning=FALSE, message=FALSE}
SubSpec_DOW <- AcuteOT %>% 
  group_by(Specialty_Description, DOW) %>% 
  summarise(Count = n())

SubSpec_DOW <- spread(SubSpec_DOW, DOW, Count)
SubSpec_DOW <- SubSpec_DOW %>% 
  filter(Specialty_Description != "Gastroenterology" & Specialty_Description != "General Medicine" & Specialty_Description != "Pain Relief")

kable(SubSpec_DOW)
```

and the top ten most prevalent procedures were;

```{r june_top_10, echo=FALSE, warning=FALSE, message=FALSE}
AcuteCase <- AcuteOT %>% 
  group_by(Operation_Description) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

kable(head(AcuteCase, n = 10))
```

Of those acute cases undertaken after hours the most frequent were;

```{r june_top_10_ah, echo=FALSE, warning=FALSE, message=FALSE}
AcuteDay <- AcuteOT %>% 
  filter(TOD == "Daytime") %>% 
  group_by(Operation_Description) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

kable(head(AcuteDay, n = 10))
```

and in the after hours acute procedures;

```{r june_top_10_day, echo=FALSE, warning=FALSE, message=FALSE}
AcuteNight <- AcuteOT %>% 
  filter(TOD == "Afterhours") %>% 
  group_by(Operation_Description) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

kable(head(AcuteNight, n = 10))
```

##November 2017
Between November and June of 2017, there had been changes in the management of acute cases across all specialties but particularly Orthopaedic surgery.

###Theatre Location
```{r nov_theatre, echo=FALSE, message=FALSE,warning=FALSE}
Nov17 <- read_csv("~/R-projects/Clinical_audits/Theatre/OTIS_Nov17.csv")

Nov17 <- Nov17 %>% 
  rename(Event_Date = 'Event Date', Operation__Description  = 'Operation Description', Speciality_Description = `Specialty Description`)

Nov17$Event_Date <-  dmy(Nov17$Event_Date)
Nov17$Theatre <- as.factor(Nov17$Theatre)

Nov17 <- Nov17 %>% 
  rename(Anaesthesia_Start = anaesthesia_start_datetime, Recovery_Takeover = recovery_takeover_datetime)

Nov17$Anaesthesia_Start <- dmy_hm(Nov17$Anaesthesia_Start)
Nov17$Recovery_Takeover <- dmy_hm(Nov17$Recovery_Takeover)


Nov17$Anaes_Span <- interval(Nov17$Anaesthesia_Start, Nov17$Recovery_Takeover)
Nov17$Anaes_Span <- as.duration(Nov17$Anaes_Span)
Nov17$Anaes_Span <- as.integer(Nov17$Anaes_Span)  / 60

Nov17 <- Nov17 %>% 
  filter(0 < Anaes_Span & Anaes_Span < 400)

Nov17$DOW <- wday(Nov17$Event_Date)
Nov17$Hour <- hour(Nov17$Anaesthesia_Start)
Nov17$Theatre <- as.factor(Nov17$Theatre)

Nov17$TOD <-  ifelse(8 <= Nov17$Hour & Nov17$Hour <= 17, "Daytime", "Afterhours" )
Nov17$TOD <- as.factor(Nov17$TOD)

OTIS17 <- Nov17

Nov17 <- Nov17 %>% 
  filter("2017-11-01" <= Event_Date & Event_Date <= "2017-11-30" & AdmissionType == "AC") %>% 
  filter(Theatre != "G", Theatre != "M", Theatre != "O", Theatre != "R", Theatre != "T")

ggplot(data = Nov17) +
  geom_bar(aes(x = Theatre, fill = Theatre)) +
  labs(title = "Number of cases per theatre Nov.2017", x = "Theatre", y = "Count of Acute cases") +
  theme_minimal()

```

Aside from a slight rise in number of acute cases occurring in theatre 3 with the corresponding reduction in utilization of theatre 4, little has changed between June and November as location where acute cases are performed.

###By day of the week
Reveals the peak incidence has moved back from Friday to Thursday as well as the total number of acute cases is less.

```{r Nov_dow, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = Nov17) +
  geom_bar(aes(x = DOW, fill = factor(DOW, labels = c("Sun", "Mon", "Tues", "Wed", "Thur", "Fri", "Sat")))) +
  labs(title = "Count by Day of Week, Nov.2017", x = "Day of week") +
  guides(fill = guide_legend(title = "Day of week")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks = element_blank())
```

and this is in contrast to the Friday peak noted in June.17.

```{r nov_comp_dow, echo=FALSE, message=FALSE, warning=FALSE}
P1 <- ggplot(data = AcuteOT) +
  geom_bar(aes(x = DOW, fill = factor(DOW, labels = c("Sun", "Mon", "Tues", "Wed", "Thur", "Fri", "Sat")))) +
  labs(title = "June.2017", x = "Day of week") +
  guides(fill = guide_legend(title = "Day of week")) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks = element_blank())
  
P2 <- ggplot(data = Nov17) +
  geom_bar(aes(x = DOW, fill = factor(DOW, labels = c("Sun", "Mon", "Tues", "Wed", "Thur", "Fri", "Sat")))) +
  labs(title = "Nov.2017", x = "Day of week") +
  guides(fill = guide_legend(title = "Day of week")) +
  ylim(0, 50) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks = element_blank())

grid.arrange(P1, P2, ncol = 2)
```


###By Sub-speciality
Reveals that Unlike June, in November Orthopaedic surgery as a sub-specialty undertook the most acute procedures. Also do note the reduction in the number of acute cases undertaken by O&G. This reduction by O&G is noticeable but not significant (P = 0.21)

```{r Nov_spec, echo=FALSE, message=FALSE, warning=FALSE}
SubSpec <- Nov17 %>% 
  group_by(Speciality_Description) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

kable(SubSpec)
```

###By surgical procedure
The type of procedures undertaken, reflect the overall utilization by sub-specialty and the type of case that require acute care.

```{r Nov_day, echo=FALSE, message=FALSE, warning=FALSE}
AcuteDay <- Nov17 %>% 
  filter(TOD == "Daytime") %>% 
  group_by(Operation__Description) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

kable(head(AcuteDay, n = 10))
```

And the top 10 type of cases requiring acute surgery after hours;

```{r Nov_night, echo=FALSE, message=FALSE, warning=FALSE}
AcuteNight <- Nov17 %>% 
  filter(TOD == "Afterhours") %>% 
  group_by(Operation__Description) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

kable(head(AcuteNight, n = 10))
```

###By hour of day
The number of acute cases performed outside of working hours has shown an obvious reduction in Nov.2017

```{r nov_how, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = Nov17, aes(x = Hour)) +
  geom_bar(aes(fill = TOD)) +
  labs(title = "Acute cases by hour of day, Nov.2017") +
  ylim(0, 25) +
  guides(fill = guide_legend(title = "Time of Day")) +
  theme_minimal()
```

In comparison to June.2017 the other notable change is that the acute cases are being performed at the end of the day as opposed to the beginning;

```{r nov_jun_hod, echo=FALSE, message=FALSE, warning=FALSE}
P1 <- ggplot(data = AcuteOT, aes(x = Hour)) +
  geom_bar(aes(fill = TOD)) +
  labs(title = "June.2017") +
  ylim(0, 25) +
  guides(fill = guide_legend(title = "Time of Day")) +
  theme_minimal() +
  theme(legend.position = "top")

P2 <- ggplot(data = Nov17, aes(x = Hour)) +
  geom_bar(aes(fill = TOD)) +
  labs(title = "Nov.2017") +
  ylim(0, 25) +
  guides(fill = guide_legend(title = "Time of Day")) +
  theme_minimal() +
  theme(legend.position = "top")

grid.arrange(P1, P2, ncol = 2)
```

###Duration
```{r Nov_duration, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(Nov17) +
  geom_density(aes(Anaes_Span, fill = TOD)) +
  labs(title = "Density plot of case duration, Nov.2017", x = "Anaesthetic duration (minutes)") +
  guides(fill = guide_legend(title = "Time of Day")) +
  theme_minimal()
```

There has been a significant increase in the mean duration of after hours acute cases (P = 0.006), so the previous significant difference in June between the duration of cases performed during the day and after hours (P < 0.001) has been eliminated (P = 0.36).

In comparison to June.2017 the density plot shows both the reduction in numbers of acute cases performed after hours and the increase in the mean duration of the cases.

```{r nov_jun_dur, echo=FALSE, message=FALSE, warning=FALSE}

P1 <- ggplot(AcuteOT) +
  geom_density(aes(Anaes_Span, fill = TOD)) +
  labs(title = "Jun.2017", x = "Anaesthetic duration (minutes)") +
  ylim(0, 0.015) +
  guides(fill = guide_legend(title = "Time of Day")) +
  theme_minimal() +
  theme(legend.position = "top")

P2 <- ggplot(Nov17) +
  geom_density(aes(Anaes_Span, fill = TOD)) +
  labs(title = "Nov.2017", x = "Anaesthetic duration (minutes)") +
  ylim(0, 0.015) +
  guides(fill = guide_legend(title = "Time of Day")) +
  theme_minimal() +
  theme(legend.position = "top")

grid.arrange(P1, P2, ncol = 2)
```

##Summary
The results from June describe how most acute cases are done in the acute theatre. The short duration cases are done at night, whilst the longer duration acute cases are done during working hours. The ratio of acute cases undertaken during the day compared to after hours is 2:1. By specialty General Surgery is the most prevalent followed by orthopedic surgery and the O&G. Cases build by over the week, making Friday the most common day for acute cases to occur, with this build up being most marked in General Surgery. 

In contrast November 2017 reveals a number of changes. The number of acute cases has reduced in general, but especially in the after hours. Notably, but not significantly, O&G has reduced its utilization of acute theatre in Nov.2017. Associated with this reduction in utilization was a statistically significant growth in the mean duration of cases. There is a slight shift back in the busiest day of acutes from Friday in June to Thursday in November. The shift is slight, not significant and is accompanied by more orthopaedic acutes in November as compared to General surgery. The number of cases performed in each theatre has remained similar.

Given the later I performed a separate analysis of "acute arranged"" to test if the reduction was a classification artificial, but this analysis did not reveal significant change between June and November. If anything it mirrored the changes seen in the Acute theatre utilization.

The most likely cause for the changes seen in this analysis is a cultural shift so that acute surgery, particular after hours, to only those cases where life or limb are under threat. 